Computing Entropies With Nested Sampling

### 2017-07-31 paper/ms.tex
More Skilling comments addressed

To overcome this, NS evolves a population of $N$ particles in the parameter
space towards higher likelihood regions.
The particles are initially initialised from the prior $\pi(\x)$, and the
particle with the lowest likelihood, $L^*_1$, is found and its details are
recorded as output. This worst particle is then discarded
and replaced by a new particle
drawn from the prior $\pi$ but subject to the constraint that its likelihood
must be above $L^*_1$ (this is usually achieved by cloning a surviving particle
and evolving it with MCMC to explore
$\pi(\x)\mathds{1}\left(L(\x) > L^*_1\right)$).
This process is repeated, resulting in an
increasing sequence of likelihoods
\begin{align}
L^*_1, L^*_2, L^*_3,... \label{eqn:likelihood_sequence}
\end{align}
Defining $X(\ell)$ as the amount of prior mass with likelihood greater than
some threshold $\ell$,
\begin{align}
X(\ell) &= \int \pi(\x) \mathds{1}\left(L(\x) > \ell\right) \, d\x,
\end{align}
each discarded point in the sequence can be assigned an $X$-value,
which transforms $Z$ to a one-dimensional integral.
\citet{skilling2006nested}'s key insight was that the
$X$-values of the discarded points are approximately known from the nature of the
algorithm. Specifically, each iteration reduces the prior mass by approximately
a factor $e^{-1/N}$. More accurately, the conditional distribution of
$X_{i}$ given $X_{i-1}$ (and the definition of the algorithm)
is obtainable by letting
$t_i \sim \textnormal{Beta}(N,1)$ and setting $X_{i+1} := t_iX_{i-1}$ (where
$X_1 = t_1$).
The $t_i$ variables represent the proportion of the remaining prior mass
that is {\em retained} at each iteration, after imposing the constraint of the
latest $L^*$ value.

@@

Consider the basic statistics problem of inferring a quantity $\mu$ from
100 observations $\boldsymbol{x} = \{x_1, ..., x_{100}\}$ whose
probability distribution (conditional on $\mu$) is
\begin{align}
p(x_i | \mu) &\sim \textnormal{Normal}(\mu, 1).
\end{align}
If the prior for $\mu$ is Normal$(0, 10^2)$, then the posterior is
\begin{align}
\mu | \boldsymbol{x} &\sim \textnormal{Normal}\left(
                                       \frac{1}{100}\sum_{i=1}^{100} x_i,
                                       \left[\frac{10}{\sqrt{10001}}\right]^2
                                       \right).
\end{align}
The true value of the differential entropy $H(\boldsymbol{x})$ is
available analytically
in this case, and its value is $146.499$ nats. However, in most situations, the
entropy of a marginal distribution is not available in closed form.

@@

Letting $\tau = \log_{10} T$, and treating
$A$ and $\phi$ are nuisance parameters, I calculated the conditional entropy
$H(\tau | \boldsymbol{Y})$ using the algorithm with a tolerance of $r = 10^{-5}$.
The result was
\begin{align}
\left<\depth\right> &= 5.379 \pm 0.038
\end{align}
Converting to a differential entropy by adding $\ln (2 \times 10^{-5})$, the conditional
entropy is
\begin{align}
H(\tau | \boldsymbol{Y}) &= -5.441 \pm 0.038 \textnormal{ nats}.
\end{align}
Since the Uniform(0,1) prior for $\tau$ has a differential entropy of zero, the
mutual information is $I(\tau; \boldsymbol{Y}) = H(\tau) - H(\tau | \boldsymbol{Y}) = 5.441 \pm 0.038$
nats.

@@

Throughout this paper, the question of how to sample from the constrained
distributions required by NS has not been discussed in detail. For the examples,
I used the original \citet{skilling2006nested} suggestion of using a short MCMC run
initialised at a surviving point. In some problems, this will not be effective,
and this will cause the resulting estimates to be incorrect, as in standard
Nested Sampling applications. Testing the sensitivity of the results to
changes in the number of MCMC steps used is a simple way to increase one's
confidence in the results, but is not a guarantee.
Despite this weakness, I hope the algorithm will be
useful in a wide range of applications.

>>>>

To overcome this, NS evolves a population of $N$ particles in the parameter
space towards higher likelihood regions.
The particles are initially initialised from the prior $\pi(\x)$, and the
particle with the lowest likelihood, $L^*_1$, is found and its details are
recorded as output. This worst particle is then discarded
and replaced by a new particle
drawn from the prior $\pi$ but subject to the constraint that its likelihood
must be above $L^*_1$ (this is usually achieved by cloning a surviving particle
and evolving it with MCMC to explore
$\pi(\x)\mathds{1}\left(L(\x) > L^*_1\right)$).
This process is repeated, resulting in an
increasing sequence of likelihoods
\begin{align}
L^*_1, L^*_2, L^*_3,... \label{eqn:likelihood_sequence}
\end{align}
Defining $X(\ell)$ as the amount of prior mass with likelihood greater than
some threshold $\ell$,
\begin{align}
X(\ell) &= \int \pi(\x) \mathds{1}\left(L(\x) > \ell\right) \, d\x,
\end{align}
each discarded point in the sequence can be assigned an $X$-value,
which transforms $Z$ to a one-dimensional integral.
\citet{skilling2006nested}'s key insight was that the
$X$-values of the discarded points are approximately known from the nature of the
algorithm. Specifically, each iteration reduces the prior mass by approximately
a factor $e^{-1/N}$. More accurately, the conditional distribution of
$X_{i}$ given $X_{i-1}$ (and the definition of the algorithm)
is obtainable by letting
$t_i \sim \textnormal{Beta}(N,1)$ and setting $X_{i+1} := t_iX_{i-1}$ (where
$X_1 = t_1$).
The $t_i$ quantities are the proportion of the remaining prior mass
that is {\em retained} at each iteration, after imposing the constraint of the
latest $L^*$ value.

@@

Consider the basic statistics problem of inferring a quantity $\mu$ from
100 observations $\boldsymbol{x} = \{x_1, ..., x_{100}\}$ whose
probability distribution (conditional on $\mu$) is
\begin{align}
p(x_i | \mu) &\sim \textnormal{Normal}(\mu, 1).
\end{align}
If the prior for $\mu$ is Normal$(0, 10^2)$, then the posterior is
\begin{align}
\mu | \boldsymbol{x} &\sim \textnormal{Normal}\left(
                                       \frac{1}{100.01}\sum_{i=1}^{100} x_i,
                                       \left[\frac{10}{\sqrt{10001}}\right]^2
                                       \right).
\end{align}
The true value of the differential entropy $H(\boldsymbol{x})$ is
available analytically
in this case, and its value is $146.499$ nats. However, in most situations, the
entropy of a marginal distribution is not available in closed form.

@@

Letting $\tau = \log_{10} T$, and treating
$A$ and $\phi$ are nuisance parameters, I calculated the conditional entropy
$H(\tau | \boldsymbol{Y})$ using the algorithm with a tolerance of $r = 10^{-5}$.
The result was
\begin{align}
\left<\depth\right> &= 5.379 \pm 0.038
\end{align}
Converting to a differential entropy by adding $\ln (2r) = -10.820$, the conditional
entropy is
\begin{align}
H(\tau | \boldsymbol{Y}) &= -5.441 \pm 0.038 \textnormal{ nats}.
\end{align}
Since the Uniform(0,1) prior for $\tau$ has a differential entropy of zero, the
mutual information is $I(\tau; \boldsymbol{Y}) = H(\tau) - H(\tau | \boldsymbol{Y}) = 5.441 \pm 0.038$
nats.

@@

Throughout this paper, the question of how to sample from the constrained
distributions required by NS has not been discussed in detail. For the examples,
I used the original \citet{skilling2006nested} suggestion of using a short MCMC run
initialised at a surviving point. In some problems, this will not be effective,
and this will cause the resulting estimates to be incorrect, as in standard
Nested Sampling applications. Testing the sensitivity of the results to
changes in the number of MCMC steps used is a simple way to increase one's
confidence in the results, but is not a guarantee (no general guarantee can
exist, because any algorithm based on sampling a subset of a space might miss
a crucial by hard-to-locate dominant spike).
Despite this weakness, I hope the algorithm will be
useful in a wide range of applications.
